Mainstreaming Scientific Racism

### 2022-11-02 talk/talk.qmd
working on slides, incld Netlify upload

---
format: 
    revealjs:
        scrollable: true
        width: 1.1*1440
        height: 1080
        toc: true
        toc-depth: 1
        toc-title: Overview
        slide-number: true
        theme: [default, custom.scss]

title: "Mainstreaming Scientific Racism"
author: "Dan Hicks & Emilio Lobato"
institute: "University of California, Merced"
bibliography: ../paper/race_science.yaml
---

# Race science in the 20th century

@@

## Before *Brown* {.smaller}

:::: columns 
::: {.column width="33%"}
![[Charles Davenport (1866-1944)](https://en.wikipedia.org/wiki/Charles_Davenport), circa 1929. [Source](https://en.wikipedia.org/wiki/File:Charles_Benedict_Davenport.jpg)](img/Charles_Benedict_Davenport.jpg){width="70%"}
:::

::: {.column width="33%"}
![[Franz Boas (1858-1942)](https://en.wikipedia.org/wiki/Franz_Boas), circa 1915. [Source](https://en.wikipedia.org/wiki/File:FranzBoas.jpg)](img/FranzBoas.jpg){width="70%"}
:::

::: {.column width="33%"}
![[Margaret Mead (1901-1978)](https://en.wikipedia.org/wiki/Margaret_Mead), 1948. [Source](https://en.wikipedia.org/wiki/File:Margaret_Mead_(1901-1978).jpg)](img/Margaret_Mead_(1901-1978).jpg){width="70%"}
:::
::::

- 1910: Davenport founds Eugenics Records Office at Cold Spring Harbor
- 1926: NRC "Committee on the Negro" includes both Davenport and Boas: "neither was influential enough to veto his rival's participation" [@BarkanRetreatScientificRacism1992 113]
- 1930s: *Coming of Age in Samoa* [@MeadComingAgeSamoa1928] popularizes cultural anthropology, "free[ing] anthropology from the shackles of biology" [@BarkanRetreatScientificRacism1992 134]
- 1940: Carnegie Institute defunds Eugenics Records Office
- 1950: "For all practical social purposes 'race' is not so much a biological phenomenon as a social myth" [@BeagleholeRaceQuestion1950; though see @BrattainRaceRacismAntiracism2007]

## The decline of eugenics and the Pioneer Fund {.smaller}

@@

Founded in 1937 to "support academic research and the \'[sic]dissemination of information, into the 'problem of heredity and eugenics' and 'the problems of race betterment'" [@MehlerFoundationFascismNew1989 21, quoting Laughlin]
<!-- 
- Presidents
	- Harry Laughlin, 1937-1941
	- Frederick Osborn, 1941–1958
	- Harry Weyher, Jr., 1958-2002
	- J. Phillippe Rushton, 2002-12
	- Richard Lynn, 2012-??
 -->

## *Brown v Board* and *Mankind Quarterly*

:::: columns
::: {.column width="30%"}
![[Reginald Ruggles Gates (1882-1962)](https://en.wikipedia.org/wiki/Reginald_Ruggles_Gates), 1921. [Source](https://en.wikipedia.org/wiki/File:Reginald_Ruggles_Gates_Portrait.jpg)](img/Reginald_Ruggles_Gates_Portrait.jpg)
:::

::: {.column width="30%"}
![[Henry Garrett (1894-1973)](https://en.wikipedia.org/wiki/Henry_Garrett_(psychologist)). [Source](https://www.psychometricsociety.org/post/past-present-and-incoming-presidents), citing American Psychologist (1946), 1(9), 371](img/henry_e_garrett.png){width="100%"}
:::

::: {.column width="30%"}
![[George Robert Gayre of Gayre and Nigg (1907-1996)](https://en.wikipedia.org/wiki/Robert_Gayre). [Source](https://www.patrickcomerford.com/2020/07/robert-gair-racist-from-rathmines-who.html), original uncredited](img/Robert Gayre profile.jpg)
:::
::::

@@

# Race science agnotology

## The Tobacco Strategy {.smallish}

Five core strategies used by industry to manufacture doubt and create the appearance of legitimate scientific controversy [after @FernandezPintoKnowBetterNot2017, esp. 59]

1. Emphasize scientific uncertainty
2. Support friendly research
3. Recruit distinguished scientists as spokesperson
4. [Construct echo chambers]{.red}
	- "creation and distribution of non-peer-reviewed journals and pamphlets"
	- "funding and organizing of symposiums" 
	- ["In this way, the tobacco industry circumvented peer review standards in publication"]{.red}
5. Attack unfavorable research

[Did Pioneer and *Mankind Quarterly* function in this way for race science?]{.red}

@@

![Journals publishing 6 or more Pioneer-funded researchers, WOS](img/wos_results.png)

@@

:::: columns
::: {.column width="60%"}
![Diachronic evolution of topic categories in *Philosophy of Science*, 1934–2015. @MalaterreWhatThisThing2019 fig. 2.](img/MalaterreWhatThisThing2019.jpeg){width="100%"}
:::

@@

- Race & intelligence research did not appear to originate in MQ and spread to other journals
- Instead, Intelligence and PID were the primary "echo chambers" used to legitimize this psychological form of race science
- "The agnotology is coming from inside the house"

@@

![Authors by topic](../out/_authors.png)

>>>>

---
format: 
    revealjs:
        scrollable: true
        width: 1.1*1440
        height: 1080
        toc: false
        toc-depth: 1
        toc-title: Overview
        slide-number: true
        theme: [default, custom.scss]

title: "Mainstreaming Scientific Racism"
author: "Dan Hicks & Emilio Lobato"
institute: "University of California, Merced<br><https://mainstream-race-science.netlify.app>"
bibliography: ../paper/race_science.yaml
---

# Introduction

## The return(?) of race science

:::: columns
::: column
![](img/Saini.png){width="70%"}\
@SainiSuperiorReturnRace2019
:::
::: column
> [Many people assumed that the eugenicists had all but disappeared with the Nazi regime, and that race science was pretty much finished at the same time.]{.blue .fragment}

> [Intellectual racism has always existed, and indeed for a chunk of history, it thrived. I believe it is still the toxic little seed at the heart of academia. However dead you might think it is, it needs only a little water, and now it’s raining.]{.red .fragment} 
:::
::::

## Institutions of race science

Pioneer Fund
: Founded 1937, during the decline of (institutionalized) eugenics

*Mankind Quarterly*
: Founded 1960 by marginalized academics as part of the backlash to *Brown v Board*

. . .

The Tobacco Strategy
: Used by regulated industries to create the appearance of legitimate scientific controversy and thereby block or delay regulation [@OreskesMerchantsDoubtHow2011; @FernandezPintoKnowBetterNot2017]

## Echo chambers

- "creation and distribution of non-peer reviewed journals and pamphlets"
- "funding and organizing of symposiums" for industry-funded research
- ["In this way, the tobacco industry circumvented peer review standards in publication"]{.red} [@FernandezPintoKnowBetterNot2017]

. . .

- "a bounded, enclosed media space that has the potential to both magnify the messages delivered within it and [insulate them from rebuttal]{.red}" [@JamiesonEchoChamberRush2008 76]

. . .

[**Did Pioneer and *Mankind Quarterly* nurture race science by protecting it from mainstream peer review?**]{}

<!-- 
# Race science, 1910-1960

@@

<!--
## Rise and fall of eugenics {.smaller}

:::: columns 
::: {.column width="33%"}
![[Charles Davenport (1866-1944)](https://en.wikipedia.org/wiki/Charles_Davenport), circa 1929. [Source](https://en.wikipedia.org/wiki/File:Charles_Benedict_Davenport.jpg)](img/Charles_Benedict_Davenport.jpg){width="80%"}
:::

::: {.column width="33%"}
![[Franz Boas (1858-1942)](https://en.wikipedia.org/wiki/Franz_Boas), circa 1915. [Source](https://en.wikipedia.org/wiki/File:FranzBoas.jpg)](img/FranzBoas.jpg){width="80%"}
:::

::: {.column width="33%"}
![[Margaret Mead (1901-1978)](https://en.wikipedia.org/wiki/Margaret_Mead), 1948. [Source](https://en.wikipedia.org/wiki/File:Margaret_Mead_(1901-1978).jpg)](img/Margaret_Mead_(1901-1978).jpg){width="80%"}
:::
::::

- 1910: Davenport founds Eugenics Records Office at Cold Spring Harbor
- 1926: NRC "Committee on the Negro" includes both Davenport and Boas: "neither was influential enough to veto his rival's participation" [@BarkanRetreatScientificRacism1992 113]
- 1930s: *Coming of Age in Samoa* [@MeadComingAgeSamoa1928] popularizes cultural anthropology, "free[ing] anthropology from the shackles of biology" [@BarkanRetreatScientificRacism1992 134]
- 1940: Carnegie Institute defunds Eugenics Records Office
<!-- - 1950: "For all practical social purposes 'race' is not so much a biological phenomenon as a social myth" [@BeagleholeRaceQuestion1950; though see @BrattainRaceRacismAntiracism2007] -->

<!--
## The decline of eugenics and the Pioneer Fund {.smaller}

@@

Founded in 1937 to "support academic research and the \'[sic]dissemination of information, into the 'problem of heredity and eugenics' and 'the problems of race betterment'" [@MehlerFoundationFascismNew1989 21, quoting Laughlin]
<!-- 
- Presidents
	- Harry Laughlin, 1937-1941
	- Frederick Osborn, 1941-1958
	- Harry Weyher, Jr., 1958-2002
	- J. Phillippe Rushton, 2002-12
	- Richard Lynn, 2012-??
 -->

<!--
## *Brown v Board* (1954) and *Mankind Quarterly*

:::: columns
::: {.column width="32%"}
![[Henry Garrett (1894-1973)](https://en.wikipedia.org/wiki/Henry_Garrett_(psychologist)). [Source](https://www.psychometricsociety.org/post/past-present-and-incoming-presidents), citing American Psychologist (1946), 1(9), 371](img/henry_e_garrett.png){width="100%"}
:::

::: {.column width="32%"}
![[Reginald Ruggles Gates (1882-1962)](https://en.wikipedia.org/wiki/Reginald_Ruggles_Gates), 1921. [Source](https://en.wikipedia.org/wiki/File:Reginald_Ruggles_Gates_Portrait.jpg)](img/Reginald_Ruggles_Gates_Portrait.jpg)
:::

::: {.column width="32%"}
![[George Robert Gayre of Gayre and Nigg (1907-1996)](https://en.wikipedia.org/wiki/Robert_Gayre). [Source](https://www.patrickcomerford.com/2020/07/robert-gair-racist-from-rathmines-who.html), original uncredited](img/Robert Gayre profile.jpg)
:::
::::

@@

-->

# Mainstreaming race science

@@

![Journals publishing 6 or more Pioneer-funded researchers, WOS author search results](img/wos_results.png)

@@

:::: columns
::: {.column width="60%"}
![Diachronic evolution of topic categories in *Philosophy of Science*, 1934-2015. @MalaterreWhatThisThing2019 fig. 2.](img/MalaterreWhatThisThing2019.jpeg){width="100%"}
:::

@@

- Race & intelligence research did not appear to originate in *MQ* and spread to other journals
- Instead, *Intelligence* and *PID* were the primary venues used to legitimize this psychological form of race science
- "The race science is coming from inside the house!"

# Thanks! {visibility="uncounted"}

<https://mainstream-race-science.netlify.app>

# Bonus content {visibility="uncounted"}

@@

![Authors by topic](img/_authors.png)
